2.4 PRNT50

### 2024-06-11 2_scripts/2.1 APL Data Quality Check_new.qmd
Finished APL ver2 data cleaning (Jan 2024)

# Second version of APL: provided by Carmen in Jan 2024

@@

```{r}
library(dplyr)         # Data management
library(lubridate)     # Manage variables in Date formats
library(zoo)           # for yearmonth Date format
library(reshape2)      # for melt() function
```

@@

## The distribution of time from NAAT to Sero-assay

@@

## Records are not patients, how many individual has been included?

@@

## Why unique PERSON_ID and ULI_ID are not matching with each other?

@@

## Scenario Examples Showcase

@@

## Add clean_IDe and Scenario back to dataset (encrypted)

@@

#### ULI ID and PERSON_ID, in NAAT+ subset

@@

## Check N=2865 cases with missing in both ULI and Person_ID

@@

## The Whole APL dataset

@@

## Categorize the time since last NAAT test

@@

## Vaccination Status

@@

Estimation on number of individuals, using cleaned version of IDe

```{r}
# Total num of individual in the data:
table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')    # N=188,722, plus 1 NA

@@

Some additional cleaning/editing, making anti-S quantitative

```{r}
# Missing values
summary(is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F) 
# N=36, beware of "not available" values, better recode those as NAs

@@

#### Yuan found this: anti-N is only available for the Omicron stage

```{r}
# The APL Data:
RFD4682_e$COLLECTION_DATE %>%
  summary()

@@

# load updated version of APL
library(readxl)
RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
    sheet = "RFD4682_Full_Data", col_types = c("text", 
        "numeric", "text", "numeric", "numeric", 
        "date", "text", "text", "date", "date", 
        "text", "numeric", "numeric", "text", 
        "text", "text", "text", "text", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric"))
View(RFD4682_v2)

@@

#### Save a cleaned ver of RFD4682e

@@

save(RFD4682_e1, file = "../1_data/private/RFD4682e1.RData")
```

>>>>

# Second version of APL: provided by Carmen in Jan 2024
# We load all columns as char, to avoid potential data corruption issues
library(readxl)
RFD4682_v2 <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
    sheet = "RFD4682_Full_Data", col_types = c("text", 
        "text", "text", "text", "text", 
        "date", "text", "text", "date", "date", 
        "text", "text", "numeric", "text", 
        "text", "text", "text", "text", "text", 
        "text", "text", "text", "text", "text", 
        "text"))
View(RFD4682_v2)

RFD4682_v2_vax <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
    sheet = "Vax Data", col_types = c("text", "numeric", "date", "text"))

RFD4682_v2_NAAT <- read_excel("../1_data/private/RFD4682_Data_for release Jan 2024.xlsx", 
    sheet = "NAAT Results", col_types = c("text", "date", "text"))

# There's some minor updates they made to the v2 (Jan 2024), so it's recommended to rerun the data cleaning code on v2, instead of adding new variables to v1 (June 2023)
```

@@

```{r}
library(dplyr)         # Data management
# library(lubridate)     # Manage variables in Date formats
library(zoo)           # for yearmonth Date format
# library(reshape2)      # for melt() function
```

@@

## 1. APL v1 (June 2023) - Data Cleaning

#### 1.1 Some Desc Analysis

##### The distribution of time from NAAT to Sero-assay

@@

##### Records are not patients, how many individual has been included?

@@

##### Why unique PERSON_ID and ULI_ID are not matching with each other?

@@

#### 1.2 Scenario Examples Showcase

@@

#### 1.3 Data cleaning

##### Add clean_IDe and Scenario back to dataset (encrypted)

@@

##### ULI ID and PERSON_ID, in NAAT+ subset

@@

##### Check N=2865 cases with missing in both ULI and Person_ID

@@

##### The Whole APL dataset

@@

##### Categorize the time since last NAAT test

@@

##### Vaccination Status

@@

Estimation on number of individuals, using cleaned version of IDe
```{r}
# Total num of individual in the data:
table(is.na(unique(RFD4682_e$clean_IDe)), useNA = 'ifany')    # N=188,722, plus 1 NA

@@

Some additional cleaning/editing, making anti-S quantitative
```{r}
# Missing values
summary(is.na(RFD4682_e$`RBD-IgGII_RESULT`)==F) 
# N=36, beware of "not available" values, better recode those as NAs

@@

#### 1.4 Other data cleaning

##### Yuan found this: anti-N is only available for the Omicron stage
**Notes:** This is an error in two ver of APL. The anti-N assay results are available and misplaced in the anti-S-RBD column.
```{r}
# The APL Data:
RFD4682_e$COLLECTION_DATE %>%
  summary()

@@

# load updated version of APL

View(RFD4682_v2)

@@

#### 1.5 Save a cleaned ver of RFD4682e

@@

save(RFD4682_e1, file = "../1_data/private/RFD4682e1.RData") # ver for matt only
```

## 2. APL v2 (Jan 2024) - Data Cleaning

#### ADD order ID and other ids -- doesn't work
```{r}
RFD4682_v2$order_ID<-RFD4682_e$order_ID

cbind.data.frame(RFD4682_e$order_ID, RFD4682_e$SERO_ID, 
                 RFD4682_v2$order_ID, RFD4682_v2$SERO_ID)
```

#### 2.1 De-duplicate
```{r}
# Does duplicated records exist?
table(duplicated(RFD4682_v2))  # no-all records are unique

# use the saved duplicated SERO_ID variables to de_duplicate:
RFD4682_v2dup<-RFD4682_v2 %>%
 .[duplicated(RFD4682_v2$SERO_ID)==T,]

# Duplicated records to clean: 
RFD4682_v2dup2<-RFD4682_v2 %>%
  .[.$SERO_ID %in% RFD4682_v2dup$SERO_ID,]
RFD4682_v2dup2$SERO_ID[9:16]<-paste0(RFD4682_v2dup2$SERO_ID[9:16],
                                     rep(c("_1", "_2"),4))

# Clean, now SERO_ID is unique
RFD4682_v2<-RFD4682_v2 %>%
  .[!(.$SERO_ID %in% RFD4682_v2dup$SERO_ID),]

RFD4682_v2<-rbind.data.frame(RFD4682_v2, RFD4682_v2dup2[c(2,4,6,8,9:16),])
rm(RFD4682_v2dup, RFD4682_v2dup2)
```
##### De-duplicate SERO_ID for APL v1
```{r}
# use the saved duplicated SERO_ID variables to de_duplicate:
RFD4682_edup<-RFD4682_e %>%
 .[duplicated(RFD4682_e$SERO_ID)==T,]

# Duplicated records to clean: 
RFD4682_edup2<-RFD4682_e %>%
  .[.$SERO_ID %in% RFD4682_edup$SERO_ID,]
RFD4682_edup2<-RFD4682_edup2 %>%
  arrange(order_ID)
RFD4682_edup2$SERO_ID[9:16]<-paste0(RFD4682_edup2$SERO_ID[9:16],
                                     rep(c("_1", "_2"),4))

# Clean, now SERO_ID is unique
RFD4682_e<-RFD4682_e %>%
  .[!(.$SERO_ID %in% RFD4682_edup$SERO_ID),]

RFD4682_e<-rbind.data.frame(RFD4682_e, RFD4682_edup2[c(2,4,6,8,9:16),])
rm(RFD4682_edup, RFD4682_edup2)
```

#### 2.2 Add ULI_IDe from v1
```{r}
library(duawranglr)    # use this to encrypt the ids

## get crosswalk and admin data files
dua_cw_file <- system.file('extdata', 'dua_cw.csv', package = 'duawranglr')
admin_file <- system.file('extdata', 'admin_data.csv', package = 'duawranglr')

## set the DUA crosswalk
set_dua_cw(dua_cw_file)
```

```{r}
# Use SERO_ID and 'RBD-IgGII_RESULT' to identify each records
RFD4682_v2<-merge(RFD4682_v2, select(RFD4682_e, SERO_ID, order_ID, PERSON_IDe, ULI_IDe,
                         clean_IDe, QUINTMAT, QUINTSOC), 
      by ="SERO_ID", all = T
      )
# ULI - ULI_IDe crosswalk
ULI_crosswalk<-select(RFD4682_v2, ULI_ID, ULI_IDe)
ULI_crosswalk<-distinct(ULI_crosswalk)

table(is.na(ULI_crosswalk$ULI_IDe))
ULI_crosswalk %>%
  .[is.na(.$ULI_IDe),]

# APL ver2 seems have solved missing value issue in the ULI ID, needs to add those
table(is.na(RFD4682_v2$ULI_ID), is.na(RFD4682_v2$ULI_IDe))

table(is.na(RFD4682_v2$PERSON_ID), is.na(RFD4682_v2$PERSON_IDe))

# encrypt new ULI_ID
set.seed(240611)
ULI_crosswalk_e <- deid_dua(ULI_crosswalk, 
                      id_col = 'ULI_ID',
                      new_id_name = 'ULI_IDe2',
                      write_crosswalk = F, id_length = 15)

ULI_crosswalk$ULI_IDe2<-ULI_crosswalk_e$ULI_IDe2

# clean the ULI_ID crosswalk file
table(is.na(ULI_crosswalk$ULI_ID))
ULI_crosswalk<-ULI_crosswalk[!is.na(ULI_crosswalk$ULI_ID),]
ULI_crosswalk<-ULI_crosswalk[-64591,]
ULI_crosswalk$ULI_IDe[is.na(ULI_crosswalk$ULI_IDe)]<-ULI_crosswalk$ULI_IDe2[is.na(ULI_crosswalk$ULI_IDe)]

# apply the cross-walk to the v2 data
df.temp1<-RFD4682_v2[!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe),]
df.temp2<-RFD4682_v2[!(!is.na(RFD4682_v2$ULI_ID) & is.na(RFD4682_v2$clean_IDe)),]

df.temp1<-merge(select(df.temp1, -c("ULI_IDe")), 
             ULI_crosswalk[,1:2],
      by="ULI_ID", all.x = T, all.y = F)
df.temp1$clean_IDe<-df.temp1$ULI_IDe

RFD4682_v2<-bind_rows(df.temp1, df.temp2)
rm(df.temp1, df.temp2, ULI_crosswalk_e)
```

```{r}
# Still, there's some missing value in clean_IDe
RFD4682_v2 %>%
  .[is.na(.$clean_IDe),]
```
#### 2.3 Clean anti-N and anti-S

##### 2.3.1 Anti-N cleaning (remove index)
```{r}
# Create an indicator for misplaced anti-N
RFD4682_v2$abb_s_index<-grepl("index", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
RFD4682_v2$abb_s_AU<-grepl("ml", RFD4682_v2$`RBD-IgGII_RESULT`, ignore.case = T)
table(RFD4682_v2$abb_s_index, RFD4682_v2$abb_s_AU, useNA = 'ifany')
# 36 samples missing abbott_s info

RFD4682_v2$misplace<-rep("anti_S missing", 214776)
RFD4682_v2$misplace[RFD4682_v2$abb_s_index==T]<-"index"
RFD4682_v2$misplace[RFD4682_v2$abb_s_AU==T]<-"AU"
table(RFD4682_v2$misplace, RFD4682_v2$abb_s_index, useNA = 'ifany')

RFD4682_v2$`N-IgG_RESULT`[RFD4682_v2$misplace=="index"]<-
  RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$misplace=="index"]

RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$misplace=="index"]<-NA

RFD4682_v2<-select(RFD4682_v2, -c("abb_s_index", "abb_s_AU"))
```

```{r}
# Remove "index" for anti-N
RFD4682_v2$`N-IgG_RESULT`<-gsub('Index', '', RFD4682_v2$`N-IgG_RESULT`)
RFD4682_v2$`N-IgG_RESULT`<-as.numeric(RFD4682_v2$`N-IgG_RESULT`)

# Anti-N interp:
summary(RFD4682_v2$`N-IgG_RESULT`[RFD4682_v2$`N-IgG_INTERP`=="Positive"]) # good
summary(RFD4682_v2$`N-IgG_RESULT`[RFD4682_v2$`N-IgG_INTERP`=="Negative"]) # good
summary(RFD4682_v2$`N-IgG_RESULT`[is.na(RFD4682_v2$`N-IgG_INTERP`)]) # needs correction

RFD4682_v2$`N-IgG_INTERP`[RFD4682_v2$`N-IgG_RESULT` < 1.4 & 
                          !is.na(RFD4682_v2$`N-IgG_RESULT`)]<-"Negative"
RFD4682_v2$`N-IgG_INTERP`[RFD4682_v2$`N-IgG_RESULT` >= 1.4 & 
                          !is.na(RFD4682_v2$`N-IgG_RESULT`)]<-"Positive"

summary(RFD4682_v2$COLLECTION_DATE[!is.na(RFD4682_v2$`N-IgG_INTERP`)])
```

##### 2.3.2 Anti-S cleaning (remove AU/ml)
```{r}
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[a-z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[A-Z]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-gsub('[/<>]', '', RFD4682_v2$`RBD-IgGII_RESULT`)
RFD4682_v2$`RBD-IgGII_RESULT`<-as.numeric(RFD4682_v2$`RBD-IgGII_RESULT`)
summary(RFD4682_v2$`RBD-IgGII_RESULT`)

# >800000 is recoded as 800000
RFD4682_v2 %>%
  .[.$`RBD-IgGII_RESULT`<800000 & 100000<.$`RBD-IgGII_RESULT` &
    !is.na(.$`RBD-IgGII_RESULT`),]

# Double check in anti_S interpretation
summary(RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$`RBD-IgGII_INTERP`=="Positive"]) # -->NA
summary(RFD4682_v2$`RBD-IgGII_RESULT`[RFD4682_v2$`RBD-IgGII_INTERP`=="Negative"]) # -->NA
summary(RFD4682_v2$`RBD-IgGII_RESULT`[is.na(RFD4682_v2$`RBD-IgGII_INTERP`)]) # good

# correct anti-S interpretations
RFD4682_v2$`RBD-IgGII_INTERP`[is.na(RFD4682_v2$`RBD-IgGII_RESULT`)]<-NA

table(RFD4682_v2$`RBD-IgGII_INTERP`, useNA = 'ifany')

```
#### 2.4 Remove unencryoted individual inditifier
```{r}
RFD4682_ev2<-select(RFD4682_v2, -c(ULI_ID, PERSON_ID, misplace))
```

#### 2.5 Save the encrypted ver of APL data
```{r}
save(RFD4682_v2, file = "../1_data/private/RFD4682_v2.RData")
save(RFD4682_ev2, file = "../1_data/private/RFD4682_ev2.RData")

save(RFD4682_ev2, file = "~/../../jmy/APL/RFD4682_ev2.RData")
```

## 3. Clean other two APL sub datasets (Vax, NAAT)

```{r}
# Add encrypted IDe to vax and NAAT data
RFD4682_v2_vaxe<-merge(RFD4682_v2_vax, select(ULI_crosswalk, ULI_ID, ULI_IDe),
      by.x="ULI", by.y="ULI_ID", all.x = T, all.y = F)
# any missing?
table(is.na(RFD4682_v2_vaxe$ULI_IDe))

# also encrypt the ULI_ID in the RFD4682_vax

list_temp<-RFD4682_v2_vaxe %>%
  .[is.na(.$ULI_IDe),]
```

### 2024-06-11 2_scripts/2.4 PRNT50.qmd
Finished APL ver2 data cleaning (Jan 2024)

```{r}
# Plot: at least 1 dose
p09v1<-tb_abbsv1_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(colour="red") +
  geom_line(aes(linetype=COI), colour="red")  +
  theme_bw() + 
  ylim(c(0,1)) +
  xlab("Time of Sample Collection") + 
  ylab("Proportion")+ 
  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
                                                      "APL-Abbott (COI=402)"))+
  theme(legend.position='top')+
  ggtitle("Stratum: at least 2 doses")
p09v1
```
##### Is 1-doser the cause of the "peak" and "bucket"?
```{r}
tb_abbsv11<-APLs %>%
  .[.$abb_s_AU==1 & .$vax %in% c("1 Dose") ,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbsv11$prop.abbs1<-tb_abbsv11$pos_abbott_s/tb_abbsv11$n
tb_abbsv11$prop.abbs2<-tb_abbsv11$pos_abbott_s2/tb_abbsv11$n

@@

```{r}
# Plot: 1 dose only
p09v11<-tb_abbsv11_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(colour="red") +
  geom_line(aes(linetype=COI), colour="red")  +
  theme_bw() + 
  ylim(c(0,1)) +
  xlab("Time of Sample Collection") + 
  ylab("Proportion")+ 
  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
                                                      "APL-Abbott (COI=402)"))+
  theme(legend.position='top')+
  ggtitle("Stratum: 1 dose only")
p09v11
```
#### Add time indicator to the figure

@@

#### 4.3.1 Data cleaning and merging
```{r}
# The imputed vaccination variable
table(CBS_vax_imputed$series_s, useNA = 'ifany')

@@

table(CBSs_AB$series_s, CBSs_AB$series_bin, useNA = 'ifany')
```
Iris algorithm cannot take unvaccinated and 1 dose only apart. In our case we can create a binary var <2 doses and >=2 doses, so we can do the same stratification with APL data.
```{r}
# re-do the melt data: no vax or 1 dose only
tb_rosv0<-CBSs_AB %>%
  .[.$series_bin == 0,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_ros= sum(pos_roche_s, na.rm = T), 
            pos_ros2= sum(pos_roche_s2, na.rm = T)
            )
tb_rosv0$prop.ros1<-tb_rosv0$pos_ros/tb_rosv0$n
tb_rosv0$prop.ros2<-tb_rosv0$pos_ros2/tb_rosv0$n

>>>>

```{r}
# Plot: at least 1 dose
p09v1<-tb_abbsv1_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(colour="red") +
  geom_line(aes(linetype=COI), colour="red")  +
  theme_bw() + 
  ylim(c(0,1)) +
  xlab("Time of Sample Collection") + 
  ylab("Proportion")+ 
  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
                                                      "APL-Abbott (COI=402)"))+
  theme(legend.position='top')+
  ggtitle("Stratum: at least 2 doses")
p09v1
```

##### Is 1-doser the cause of the "peak" and "bucket"?

```{r}
tb_abbsv11<-APLs %>%
  .[.$abb_s_AU==1 & .$vax %in% c("1 Dose") ,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_abbott_s= sum(pos_abbott_s, na.rm = T), 
            pos_abbott_s2= sum(pos_abbott_s2, na.rm = T)
            )
tb_abbsv11$prop.abbs1<-tb_abbsv11$pos_abbott_s/tb_abbsv11$n
tb_abbsv11$prop.abbs2<-tb_abbsv11$pos_abbott_s2/tb_abbsv11$n

@@

```{r}
# Plot: 1 dose only
p09v11<-tb_abbsv11_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=COI, colour=COI)) +
  geom_point(colour="red") +
  geom_line(aes(linetype=COI), colour="red")  +
  theme_bw() + 
  ylim(c(0,1)) +
  xlab("Time of Sample Collection") + 
  ylab("Proportion")+ 
  scale_linetype_manual("", values=c(1,3), labels = c("APL-Abbott (COI=50)", 
                                                      "APL-Abbott (COI=402)"))+
  theme(legend.position='top')+
  ggtitle("Stratum: 1 dose only")
p09v11
```

#### Add time indicator to the figure

@@

#### 4.3.1 Data cleaning and merging

```{r}
# The imputed vaccination variable
table(CBS_vax_imputed$series_s, useNA = 'ifany')

@@

table(CBSs_AB$series_s, CBSs_AB$series_bin, useNA = 'ifany')
```

Iris algorithm cannot take unvaccinated and 1 dose only apart. In our case we can create a binary var \<2 doses and \>=2 doses, so we can do the same stratification with APL data.

```{r}
# re-do the melt data: no vax or 1 dose only
tb_rosv0<-CBSs_AB %>%
  .[.$series_bin == 0,] %>%
  group_by(samplemonth) %>%
  summarise(n = n() , 
            pos_ros= sum(pos_roche_s, na.rm = T), 
            pos_ros2= sum(pos_roche_s2, na.rm = T)
            )
tb_rosv0$prop.ros1<-tb_rosv0$pos_ros/tb_rosv0$n
tb_rosv0$prop.ros2<-tb_rosv0$pos_ros2/tb_rosv0$n
